---
title: "Quiz 3"
author: "mail@darrenbishop.com"
date: "25 February 2016"
output: html_document
---

## Question 1

```{r}
cars.lm1 <- lm(data = mtcars, mpg ~ factor(cyl) + wt)
coef(cars.lm1)[3]
```

***Answer:*** -6.07086

## Question 2

```{r}
cars.lm2 = lm(data = mtcars, mpg ~ factor(cyl))
cars.lm2w = lm(data = mtcars, mpg ~ factor(cyl) + wt)
coef(cars.lm2)[3]
coef(cars.lm2w)[3]
```

***Answer:*** Holding weight constant, cylinder appears to have less of an impact on mpg than if weight is disregarded.  

## Question 3

```{r}
fit1 = lm(data = mtcars, mpg ~ factor(cyl) + wt)

fit2 = update(fit1, mpg ~ factor(cyl) + wt + factor(cyl) * wt)

summary(anova(fit1, fit2))
```

***Answer:*** P-value is larger than a 0.05 Type-I Error rate, thus the interaction term is not necessary

## Question 4

```{r}
cars.lm4 <- lm(mpg ~ I(wt) + factor(cyl), data = mtcars)
summary(cars.lm4)$coef
```

***Answer:*** The estimated expected change in MPG per ton increase in weight for a specific number of cylinders (4, 6, 8).

## Question 5

```{r}
x = c(0.586, 0.166, -0.042, -0.614, 11.72)
y = c(0.549, -0.026, -0.127, -0.751, 1.344)

hatvalues(lm(y ~ x))
```

***Answer:*** 0.9946

## Question 6

```{r}
x = c(0.586, 0.166, -0.042, -0.614, 11.72)
y = c(0.549, -0.026, -0.127, -0.751, 1.344)

dfbetas(lm(y ~ x))
```

***Answer:*** -134

## Question 7

***Answer:*** It is possible for the coefficient to reverse sign after adjustment. For example, it can be strongly significant and positive before adjustment and strongly significant and negative after adjustment.
