---
title: "Quiz 3"
author: "mail@darrenbishop.com"
date: "25 February 2016"
output: html_document
---

## Question 1

```{r}
mtcars$cyl = factor(mtcars$cyl)

coef(lm(data = mtcars, mpg ~ cyl + wt))[3]
```

***Answer:*** -6.07086

## Question 2

```{r}
coef(lm(data = mtcars, mpg ~ cyl))[3]
```

***Answer:*** The model is biased towards cyl, that is, adjusting and holding weight constant shows cyl has less impact coefficient is closer to zero.  

## Question 3

```{r}
fit1 = lm(data = mtcars, mpg ~ cyl + wt)

fit2 = update(fit1, mpg ~ cyl + wt + cyl * wt)

summary(anova(fit1, fit2))
```

***Answer:*** P-value is larger than a 0.05 Type-I Error rate, thus the interaction term is not necessary

## Question 4

```{r}
summary(lm(mpg ~ wt + cyl, data = mtcars))

summary(lm(mpg ~ I(wt * 0.5) + cyl, data = mtcars))
```

***Answer:*** The estimated expected change in MPG per half ton increase in weight for a specific number of cylinders (4, 6, 8).

## Question 5

```{r}
x = c(0.586, 0.166, -0.042, -0.614, 11.72)
y = c(0.549, -0.026, -0.127, -0.751, 1.344)

hatvalues(lm(y ~ x))
```

***Answer:*** 0.9946

## Question 6

```{r}
x = c(0.586, 0.166, -0.042, -0.614, 11.72)
y = c(0.549, -0.026, -0.127, -0.751, 1.344)

dfbetas(lm(y ~ x))
```

***Answer:*** -134

## Question 7

***Answer:*** It is possible for the coefficient to reverse sign after adjustment. For example, it can be strongly significant and positive before adjustment and strongly significant and negative after adjustment.
