---
title: "Analysis of Automatic vs. Manual Transmission Affect on MPG"
author: "mail@darrenbishop.com"
date: "28 February 2016"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
geometry: margin=1cm
mainfont: Arial
mathfont: Courier
header-includes:
  - \usepackage{float}
  - \let\origfigure\figure
  - \let\endorigfigure\endfigure
  - \renewenvironment{figure}{\expandafter\origfigure\expandafter[H]}{\endorigfigure}
---

```{r Set Options, Libs and Data, message=FALSE, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(car)

opts_chunk$set(fig.path="figures/", fig.width=10, fig.lp = "Figure: ", fig.align = "center")
opts_chunk$set(message=FALSE, cache=TRUE)
```

## Executive Summary

This analysis aimed to answer the question of which transmission, automatic or manual, is better for mpg performance of cars, quantifying and validating any difference.  

The analysis suggests the choice of automatic vs. manual transmission has ___no significant___ affect on mpg performance.  

Despite a ___0.15 mpg difference in favour of manual-transmission___, the ___`r sprintf("%.4g to %.4g", -2.517734, 2.817941)` confidence interval is centred at 0-zero___, strongly suggesting there is no effect.  

Two significant covariates were identified, both having confidence interval wholly below 0-zero:

Weight
: A continuous variable; a one unit change representing 1000 lbs
: Per unit increase, MPG performance is reduced by 3.15 mpg

No. of Cylinders
: A discrete numeric variable; values of 4, 6 and 8 cylinders
: A 6 cylinder engine reduces MPG performace by 4.23 mpg, compared to a 4 cylinder engine
: An 8 cylinder engine reduces MPG performace by 6.08 mpg, compared to a 4 cylinder engine
: An 8 cylinder engine reduces MPG performace by 1.82 mpg, compared to a 6 cylinder engine

___If buying from a similar set of cars i.e. having the same no. of cylinders, pick the lighter one to get better MPG performance!___  

## Exploratory Data Analysis

The documentation for the `mtcars` dataset (see Aappendix) reveals that the `vs` and `am` variables are binary and `cyl`, `gear` and `carb` variables are discrete; they can all thus be represented as factor variables.  

```{r, echo = F}
data(mtcars)

cars.dt = tbl_dt(mtcars) %>%
    mutate(
        car = rownames(mtcars),
        trans = factor(am, labels = c("Automatic", "Manual")),
        engine = factor(vs, labels = c("V-Shape", "Straight")),
        cyl = factor(cyl),
        carb = factor(carb)) %>%
    dplyr::select(car, mpg, trans, engine, mpg:carb, -am, -vs)
```

`vs` and `am` are re-labelled and also renamed to `engine` and `trans`, respectively; Table 2. of the Appedix shows a subset of the `mtcars` data with engine (`engine`) and transmission (`trans`) presented with self-describing levels.  

Figure 1. of the Appendix shows a preliminary analysis, capturing the distribution of cars' `mpg` performance, grouped by transmission.  

It appears that more than 75% of automatic cars get less than 20 mpg, while more than 75% of manual cars perform better than 20 mpg, suggesting manual cars are better for fuel efficiency.  

## Model Design

Considering how cars are constructed and what they are used for, the available variables can be placed into two groups, _performance metrics_ and _mechanical properties_, which in statistics terms translates to _response_ and _regressor_ variables, respectively.  

Performance Metrics
: `mpg`, `hp` and `qsec`
: Intuition: these are derived, that is, they can not be changed directly, but indirectly through changes to the _mechanical properties_

Mechanical Properties
: `cyl`, `disp`, `drat`, `wt`, `engine`, `trans`, `gear` and `carb`
: Intuition: these are structural aspects to be built into a car, analogous to the no. of bedrooms built into a house

Any correlation between _performance metrics_ would likely be correlated and confounded by one of the _mechanical properties_.

Thus for the sake of parsimony and simplified design, this analysis disregards all _performance metrics_ when considering regressor selection.  

### ANOVA Testing

Table 3. of the Appendix shows the results of running the _mechanical properties_ through a step-wise-inclusion Analysis of Variance test; 0.05 Type-I Error rate is used for determining significance:

Model 1.
: contains the regressor of interest, transmission (`trans`)

Model 2.
: introduces Weight (`wt`) with a significant P-value of 0.0000012
: We reject the Null-Hypothesis that Weight does not explain any variance in the model

Model 4.
: introduces No. of Cylinders (`cyl`) with a significant P-value of 0.0468712
: We reject the Null-Hypothesis that No. of Cylinders does not explain any variance in the model

All interim and subsequent inclusions produced large P-values, suggesting including them would result in an over-fitted model.  

```{r, echo = F}
cars.model = lm(mpg ~ trans + wt + cyl, data = cars.dt)
```

With so few regressor variables as candidates for inclusion, there is the risk of producing a biased model. Variance Inflation Factor analysis and Regression Diagnostics are used to address this.  

### Variance Inflation Factor Analysis

Tables 4. and 5. of the Appendix shows the result of VIF anaylsis, where all 8 _mechanical properties_ and just the 3 candidate variables (`trans`, `wt` and `cyl`) are included, respectively.  

Table 4. shows large standard error inflation across the board, suggesting the presence of strongly correlated regressors; Table 5. shows much diminished standard error inflation in comparison for `trans`, `wt` and `cyl`.  

### Regression Diagnostics

To support the selection of `trans`, `wt` and `cyl` as regressor variables, Figure 2. of the Appendix gives a: Residual plot showing the abscence of pattern, which would otherwise indicate a poor model fit; and a Q-Q plot showing the standardized residuals are approximately Normal with mean zero, meaning there is no remaining variance still to be accounted for (by some excluded confounder).  

## Data Interpretation

With these _ANOVA Test_ findings, validated by the _Variance Inflation Factor Analysis_ and _Regression Diagnostics_, the final model is:

$$ Model: mpg \sim trans + wt + cyl $$

```{r, echo = F}
cars.stats = data.frame(coef(cars.model)) %>%
    rename(Coefficients = coef.cars.model.) %>%
    cbind(confint(cars.model))

kable(cars.stats, caption = "Coefficients and 95% Confidence Intervals")
```

Knowing that both `trans` and `cyl` are factors of 2 and 3 levels each, respectively, 4 parallel regression lines are expected i.e. 4 different intercepts, all with the same slope determined by the coefficient of `wt`; note that the first represents automatic with 4 cylinders.    

Table 1. shows a model intercept of 33.7535920, which as mentioned, represents cars with automatic transmission and 4 cylinder engines.  

___According to this model, the difference in switching from an automatic to a manual transission car, also with a 4 cylinder engine, is merely an extra 0.15 mpg___

While keeping weight and transmission fixed, going for a bigger engine i.e. 6 cylinders or 8 cylinders, reduces mpg performance by 4.23 and 6.08 each, respectively.  

It is intuitive that, while all other variables are kept the same, a heavier vehicle needs more energy/power to move, thus burning more fuel; this model indicates that per 1000 lbs increase in weight, the mpg performance is reduced by 3.15.  

## Conclusions

This analysis shows the choice of automatic vs. manual is somewhat moot, where the 95% confidence interval contains 0-zero i.e. likely no effect on mpg.  

This finding contradicts the preliminary analysis, which suggested that manual cars exhibited far superior mpg performance.  

The deeper analysis shows there are stronger covariates than transmission e.g. weight and # of cylinders, that better explain the variance in mpg.  

```{r, results = 'asis', echo = FALSE}
cat("\\pagebreak")
```

## Appendix

```{r, include = F}
indicate.significance <- Vectorize(function(pvalue) {
    if (is.na(pvalue)) ""
    else if (pvalue <= 0.001) "***"
    else if (pvalue <= 0.01) "**"
    else if (pvalue <= 0.05) "*"
    else if (pvalue <= 0.1) "."
    else ""
})

vif.table <- function (model) {
  model.vif = rbind(
      t(vif(model)[,1]),
      t(sqrt(vif(model))[,1])
  )
  rownames(model.vif) = c("\u03c3\u00B2", "\u03c3")
  kable(model.vif, caption = "Variance and Standard Deviation Inflation Factors")
}
```

### `mtcars` Documentation Excerpt

    A data frame with 32 observations on 11 variables.

    [, 1]	mpg	Miles/(US) gallon
    [, 2]	cyl	Number of cylinders
    [, 3]	disp	Displacement (cu.in.)
    [, 4]	hp	Gross horsepower
    [, 5]	drat	Rear axle ratio
    [, 6]	wt	Weight (1000 lbs)
    [, 7]	qsec	1/4 mile time
    [, 8]	vs	V/S
    [, 9]	am	Transmission (0 = automatic, 1 = manual)
    [,10]	gear	Number of forward gears
    [,11]	carb	Number of carburetors

_V/S_ (`vs`) is taken to mean _V-Shape_ or _Straight_ engine; to be consistent with Transmission (`am`), 'V-Shape' is assumed to be 0 and 'Straight' as 1.  

### Post-processed `mtcars` Sample

```{r, echo = F}
kable(head(cars.dt),
      caption = "`mtcars` sample, with renamed and relabelled engine and trans.")
```

### Exploratory Analysis Plots

```{r, fig.cap = "A Violin plot enriched with mean, median and inter-quartile range for transmission", echo = F}
ggplot(data = cars.dt, aes(x = trans, y = mpg, fill = trans)) +
    geom_violin(alpha = 0.5) +
    geom_boxplot(width = 0.05, aes(fill = NULL)) +
    stat_summary(fun.y = mean, geom = "point", shape = 23, size = 2) +
    scale_fill_discrete(guide = FALSE)
```

### Analysis of Variance Test Table

```{r, echo = F}
cars.model1 = lm(mpg ~ trans, data = cars.dt)
cars.model2 = lm(mpg ~ trans + wt, data = cars.dt)
cars.model3 = lm(mpg ~ trans + wt + disp, data = cars.dt)
cars.model4 = lm(mpg ~ trans + wt + disp + cyl, data = cars.dt)
cars.model5 = lm(mpg ~ trans + wt + disp + cyl + carb, data = cars.dt)
cars.model6 = lm(mpg ~ trans + wt + disp + cyl + carb + engine, data = cars.dt)
cars.model7 = lm(mpg ~ trans + wt + disp + cyl + carb + engine + gear, data = cars.dt)
cars.model8 = lm(mpg ~ trans + wt + disp + cyl + carb + engine + gear + drat, data = cars.dt)
cars.models = list(
    cars.model1, cars.model2, cars.model3, cars.model4,
    cars.model5, cars.model6, cars.model7, cars.model8
)

cars.anova = do.call(anova, cars.models)
cars.anova.heading <- strsplit(split = "\n", attr(cars.anova, "heading")[2])
cars.anova.models = data.frame(cars.anova.heading) %>%
    setNames("Formula") %>%
    add_rownames(var = "Model") %>%
    cbind(cars.anova) %>%
    mutate(
        Formula = sub("Model \\d+: ", "", Formula, perl = T),
        " " = indicate.significance(`Pr(>F)`)
    )
kable(cbind(cars.anova.models), caption = "ANOVA Test of the 8 _Mechanical Properties_")
```

### Variance Inflation Factor Analysis Tables

```{r, echo = F}
vif.table(lm(mpg ~ trans + wt + cyl + disp + drat + engine + gear + carb, data = cars.dt))
vif.table(lm(mpg ~ trans + wt + cyl, data = cars.dt))
```

### Regression Diagnostic Plots

```{r, fig.cap = "Diagnostic plots showing the absence of pattern in and normality of the residuals", echo = F}
diag.resid = ggplot(data = cars.model, aes(x = .fitted, y = .resid)) +
    geom_hline(yintercept = 0, col = "grey", linetype = "dashed") +
    stat_smooth(col = "red", alpha = 0.5) +
    geom_point() +
    labs(title = "Residuals vs. Fitted (MPG)",
         x = "Fitted (MPG)",
         y = "Residuals")

diag.qq = ggplot(data = cars.model, aes(sample = .stdresid)) +
    geom_qq() +
    geom_hline(yintercept = 0, col = "grey", linetype = "dashed") +
    geom_abline(col = "blue", linetype = "dotted") +
    labs(title = "Normal Q-Q",
         x = "Theoretical Quantiles",
         y = "Standardized Residuals")

grid.arrange(diag.resid, diag.qq, ncol = 2)
```
